Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially Dependent Interaction As a Determinant of Neural Activity and Plasticity \end{large}

### 2017-03-15 draft.tex
.

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
\usepackage[a4paper,margin=1.5in]{geometry}
%\usepackage{geometry}
%\geometry{
%a4paper,
%left=2.0in,
%right=1.0in}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{tabu}
\usepackage{float}

\usepackage[titletoc]{appendix}
\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}

@@

\title{Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\
\begin{large}
Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity
\end{large}
}

@@

\begin{titlepage}
\clearpage
\thispagestyle{empty}
\begin{center}
\LARGE{{\scshape Faculty 13 Physics}\\
  {\scshape Frankfurt Institute for Advanced Studies}\\
  \vspace{1cm}
  Master's Thesis\\}
  \vspace{1cm}
  \huge{Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network}\\
  \vspace{0.5cm}
  \LARGE{Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity}\\
  \vspace{2cm}
  \LARGE{Fabian Schubert\\
  Matriculation Number: 4282141}\\
  \vspace{2cm}
  Thesis Supervisor: Prof. Dr. Jochen Triesch\\
  \end{center}

@@

\begin{abstract}
Building upon on a self-organizing spiking neural network (LIF-SORN), we attempted to replace the intrinsic homeostatic control system used in earlier versions by a mechanism based on the diffusion of a neurotransmitter across the nervous tissue. The model of diffusive homeostasis was adopted from a paper by Sweeney et al. and models the tissue as a surface of square shape and a set of points on this surface, representing the positions of the neurons within the network. Excitatory neurons acted as a source of nitric oxide (NO), as well as as a sensor for the NO concentration at each individual position. Production and sensing of NO formed the basis of a feedback loop: Individual NO readouts caused an appropriate change within the intrinsic excitability of the neurons. The control system was closed by linking the rate of NO production to the neurons' firing rates.

The main goal of the modification of homeostasis was to allow for a broad and heavy tailed distribution of firing rates among the excitatory neural population. This feature of cortical activity has been extensively observed and studied experimentally, but could not be achieved by the formerly used single-cell homeostatic mechanism which bound firing rates of all neurons to a fixed target value. Indeed, our statistical analysis of spiking activity showed that diffusive homeostasis could significantly broaden the distribution of firing rates.

After having analyzed the stability of the homeostatic control loop and the resulting statistical features of spiking activity, we compared both homeostatic mechanisms with respect to features of synaptic network structures emerging throughout the simulation. Apart from the preservation of a number of topological features we found that diffusive homeostasis allowed for the emergence of highly influential neurons with strong outgoing synaptic efficacies. By means of an analytic approach to the homeostatic steady state, we could relate this feature of synaptic topology to the imposed spatial structure of the neural population.
\end{abstract}

@@

\section*{Glossary}
\bigskip
\begin{minipage}{0.33\textwidth} 
$\rm Ca^{2+}$\\
CDF\\
CF\\
CNS\\
CV\\
IP\\
ISI\\
$\rm K^{+}$\\
LIF\\
LTD\\
LTP\\
L5\\
MLE\\
nNOS\\
NO\\
PDF\\
SORN\\
STD\\
STDP\\
STF\\
STP
\end{minipage}
\hfill
\begin{minipage}{0.66\textwidth}
ionic calcium\\
cumulative distribution function\\
connection fraction\\
central nervous system\\
coefficient of variation\\
intrinsic plasticity\\
interspike interval\\
ionic potassium\\
leaky-integrate-and-fire\\
long-term depression\\
long-term potentiation\\
layer five of the mammalian neocortex\\
maximum likelihood estimation\\
neural nitric oxide synthesis\\
nitric oxide\\
probability density function\\
self-organizing recurrent neural network\\
short-term depression\\
spike-timing-dependent plasticity\\
short-term facilitation\\
short-term plasticity
\end{minipage}

@@

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional roles of the mammalian central nervous system it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found so far and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons - being more or less active than a certain set point of mean activity - are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is taking place in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. One can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, Hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing intrinsic properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}.

Theoretical models of homeostasis are generally based on a form of feedback control. This is also known as a \textit{closed loop control} in control theory. All these systems include a number of necessary functional components, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.

\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
A closed loop control generally must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. It might not be possible however to directly quantify this desired output in terms of some internal state of the process. The firing rate of a neuron for example is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. The (potentially transformed) output is fed into a \textit{comparator} in the next step, which---unsurprisingly---compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. The term \textit{homeostasis} however generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. This is also the point in the control loop where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption[Schematic illustration of a closed loop control]{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal one can classify control loops by three types:
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. In the study of nonlinear feedback systems they thus only apply in a limited regime where a first order approximation of the system is sufficiently exact.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
A key step in characterizing a control loop in the context of biological systems should be to identify the previously described abstract elements. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and ``straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. Firing rate is---as stated earlier---an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a ``sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological quantities to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control, and rather points to an integral or proportional control. It is clear however that the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. The conductance of an ion channel for example has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

\subsubsection{Single-Cell Homeostasis as an Integral Control}
Former versions of the LIF-SORN and SORN used a simple form of integral control to regulate excitatory firing rates. Individual firing rates of excitatory cells were estimated by a spike counter over short subsequent time periods and compared to a target value. This difference was then fed to a controller which increased or decreased neuronal excitability in an additive way (making it a discrete form of integral control). As mentioned before, this type of control allowed to reliably tune each cell to a given target firing rate under changing input. A global target firing rate was chosen for all excitatory neurons. Consequently, this resulted in a very narrow distribution of excitatory firing rates.

\subsubsection{The Role of Firing Rate Heterogeneity---a Dichotomy between Heterogeneity and Control?}\label{Section_Role_Fir_Rate_Het}
Modeling homeostasis necessarily includes a form of error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. Of course this raises the question of how to choose these homeostatic targets. As mentioned in the previous section, former versions of the SORN defined a single global target value. However, experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. Many statistical analyses found particular evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, claim that it is reasonable to define a single global target of activity.

However, numerous experimental and theoretical studies have suggested that the presence of both slow- and fast-firing cells is not to be regarded as an ignorable, mere side effect of brain dynamics \cite{Buzsaki_2004,Marsat_2010,Tripathy_2013}. It rather seems to be that skewed and heavy-tailed firing rate statistics are necessary with respect to the preservation of stable synaptic structures, being related to long-term memory, while still allowing synaptic rewiring in order to adapt to changes in external stimuli. For example, experiments by Dragoi et al. have found that during remapping of hippocampal place cells, neurons with low firing rates are subject to stronger changes within their field of spatial representation than those with high firing rates \cite{Dragoi_2003}. In this context Buzsáki and Mizuseki argue that a small subset of highly active neurons may serve to form relatively strongly connected and stable sub-networks that can quickly react to momentary changes in the environment, while fine tuned adaptation involves the slower firing majority of cells \cite{Buzsaki_Fir_Rates_2014}. A realistic network model of neural activity and plasticity should therefore allow for a broad distribution of firing rates to cope with these functional aspects.

@@

\subsubsection{Nitric Oxide in the Nervous System}\label{NO_Experiments_Section}
The model presented in \cite{Sweeney_Paper} includes spatial interaction across excitatory neurons through the diffusion of nitric oxide (NO). Enzymes that are responsible for NO synthesis (NOS) are present in different areas of the body, namely being involved in the dilation of blood vessels (endothelial NOS, eNOS), the immune system (inducible NOS, iNOS) and the CNS (neuronal NOS, nNOS) \cite{NOS_Mammals}. These Enzymes differ in their functionality and dependence on the presence of other molecules. In particular, nNOS is sensitive to the concentration of $\mathrm{Ca^{2+}}$ ionic Calcium \cite{Knowles_Ca_nNOS,Steinert_NO}. Nitric oxide that was synthesized in the nervous system can diffuse freely because of its small molecular structure and electrical neutrality, which makes it a good candidate to be involved into a form of diffusive intercellular communication \cite{Lancaster_1994}.

@@

\subsection{Previous Research on the used Network Model}
This thesis is based on earlier versions of a self-organizing recurrent neural network (SORN), in particular an implementation using leaky integrate-and-fire neurons (LIF-SORN) \cite{SORN_Paper}. It was developed to model features of activity and topology of cortical networks, in particular L5 of the neocortex. A slow homeostatic mechanism was set up among excitatory neurons through an intrinsic adaptation of excitability in these previous implementations. This led to the aforementioned problems of homeostatic control: Neural activity was tuned to a globally set target, allowing for no variability. Despite this limitation, the model has proven to be capable of showing a number of experimentally confirmed features. Zheng et al. have shown that distribution and dynamics of synaptic efficacies measured in rat hippocampus can be reproduced by a binary SORN using a discretized version of spike-timing-dependent plasticity and presynaptic normalization \cite{Pengsheng_2013}. Miner et al. designed the LIF-SORN based on this model. This network reproduced the results of the binary SORN regarding self-organizing synaptic dynamics. Furthermore, it was able to explain the experimentally observed overrepresentation of bidirectional connections (see \cite{Markram_Connections_1997,Song_Connectivity_2005}) by means of a distance-dependent implementation of synaptic growth.

@@

\subsection{Thesis Outline}
Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neural activity which has previously been suppressed by single-cell homeostasis. Furthermore, we predict excitatory activity on the level of individual cells by an analytic approach, thereby gaining an understanding of the relation between spatial structure and firing rates. In the face of possible new features within the network's structure, we clarify the causal relation between diffusive spatial interaction and synaptic topology.

@@

The Neural Network was simulated with the code used in \cite{SORN_Paper}, which makes use of the BRIAN spiking neural network simulator \cite{Briansim}. All following explanations regarding the simulation of neurons and mechanisms of synaptic plasticity are thus based on the methods described in the aforementioned paper.

400 excitatory LIF neurons and 80 inhibitory LIF neurons were assigned random positions across a square area of 1000 $\times$ 1000 $\mu m$. All but recurrent excitatory synapses were randomly generated before the start of the simulation until a desired connection fraction was reached. The connection probability between two neurons was calculated from a distance-dependent Gaussian function with a standard deviation of 200 $\mu m$. For excitatory to inhibitory (EI) and inhibitory to excitatory (IE) synapses, the connection fraction was set to 0.1, and 0.5 for recurrent inhibitory synapses (II). These connections were kept at a fixed connection strength throughout the simulation. Furthermore, all synapses were simulated with a fixed (distance independent) conduction delay. See table \ref{syn_conn_params} for a summary of parameters.

@@

\begin{table}
\caption{Parameters of synaptic connections.}
\begin{tabu}{|l|l|l|l|l|}
\hline
\textbf{parameter} & \textbf{EE} & \textbf{EI} & \textbf{IE} & \textbf{II} \\ \hline
connection fraction & $\rightarrow 0.1$ & $0.1$ & $0.1$ & $0.5$ \\ \hline
initial connection strength & $0.0001 mV$ & $1.5 mV$ & $-1.5 mV$ & $-1.5 mV$ \\ \hline
conduction delay & $1.5 ms$ & $0.5 ms$ & $1.0 ms$ & $1.0 ms$ \\
\hline
\end{tabu}
\label{syn_conn_params}
\end{table}

>>>>

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
\usepackage[a4paper,margin=1.5in]{geometry}
%\usepackage{geometry}
%\geometry{
%a4paper,
%left=2.0in,
%right=1.0in}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{tabu}
\usepackage{float}
\usepackage{upgreek}
\usepackage[titletoc]{appendix}
\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}

@@

\title{Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\
\begin{large}
Spatially Dependent Interaction As a Determinant of Neural Activity and Plasticity
\end{large}
}

@@

\begin{titlepage}
\clearpage
\thispagestyle{empty}
\begin{center}
\LARGE{{\scshape Faculty 13 Physics}\\
  {\scshape Frankfurt Institute for Advanced Studies}\\
  \vspace{1cm}
  Master's Thesis\\}
  \vspace{1cm}
  \huge{Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network}\\
  \vspace{0.5cm}
  \LARGE{Spatially Dependent Interaction As a Determinant of Neural Activity and Plasticity}\\
  \vspace{2cm}
  \LARGE{Fabian Schubert\\
  Matriculation Number: 4282141}\\
  \vspace{2cm}
  Thesis Supervisor: Prof. Dr. Jochen Triesch\\
  \end{center}

@@

\begin{abstract}
Building on a Self-Organizing Spiking Neural Network (LIF-SORN), we attempted to replace the intrinsic homeostatic control system used in earlier versions by a mechanism based on the diffusion of a neurotransmitter across the nervous tissue. The model of diffusive homeostasis was adopted from a study by Sweeney et al. \cite{Sweeney_Paper} and models the tissue as a surface of square shape and a set of points on this surface, representing the positions of the neurons within the network. Excitatory neurons act as a source of nitric oxide (NO) as well as as a sensor for the NO concentration at each individual position. Production and sensing of NO forms the basis of a feedback loop: Individual NO readouts cause an appropriate change within the intrinsic excitability of the neurons. The control system is closed by linking the rate of NO production to the neurons' firing rates.

The main goal of the modification of homeostasis was to allow for a broad and heavy tailed distribution of firing rates among the excitatory neuronal population. This feature of cortical activity has been extensively observed and studied experimentally, but could not be achieved by the formerly used single-cell homeostatic mechanism which bound firing rates of all neurons to a fixed target value. Indeed, our statistical analysis of spiking activity showed that diffusive homeostasis could significantly broaden the distribution of firing rates.

After having analyzed the stability of the homeostatic control loop and the resulting statistical features of spiking activity, we compared both homeostatic mechanisms with respect to features of synaptic network structures emerging throughout the simulation. Apart from the preservation of a number of topological features we found that diffusive homeostasis allowed for the emergence of highly influential neurons with strong outgoing synaptic efficacies. By means of an analytic approach to the homeostatic steady state, we could relate this feature of synaptic topology to the imposed spatial structure of the neuronal population.
\end{abstract}

@@

\section*{Glossary}
\bigskip
\begin{minipage}{0.33\textwidth} 
$\rm Ca^{2+}$\\
CDF\\
CF\\
CNS\\
CV\\
IP\\
ISI\\
$\rm K^{+}$\\
LIF\\
LTD\\
LTP\\
L5\\
MLE\\
nNOS\\
NO\\
PDF\\
SORN\\
STD\\
STDP\\
STF\\
STP
\end{minipage}
\hfill
\begin{minipage}{0.66\textwidth}
ionic calcium\\
cumulative distribution function\\
connection fraction\\
central nervous system\\
coefficient of variation\\
intrinsic plasticity\\
interspike interval\\
ionic potassium\\
leaky-integrate-and-fire\\
long-term depression\\
long-term potentiation\\
layer five of the mammalian neocortex\\
maximum likelihood estimation\\
neural nitric oxide synthesis\\
nitric oxide\\
probability density function\\
Self-Organizing Recurrent Neural Network\\
short-term depression\\
spike-timing-dependent plasticity\\
short-term facilitation\\
short-term plasticity
\end{minipage}

@@

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional roles of the mammalian central nervous system it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} embracing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but for neuronal activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found so far and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons - being more or less active than a certain set point of mean activity - are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is taking place in a multiplicative manner, thereby preserving proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. One can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, Hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing intrinsic properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}.

Theoretical models of homeostasis are generally based on a form of feedback control. This is also known as a \textit{closed loop control} in control theory. All these systems include a number of necessary functional components, which will be briefly introduced in the following section. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.

\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
A closed loop control generally must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is simply to generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to regulate by means of the feedback loop and thus represents the signal that is further processed. It might not be possible however to directly quantify this desired output in terms of some internal state of the process. The firing rate of a neuron for example is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. The (potentially transformed) output is fed into a \textit{comparator} in the next step, which---unsurprisingly---compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. The term \textit{homeostasis} however generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. This is also the point in the control loop where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption[Schematic illustration of a closed loop control]{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process one can classify control loops by three types:
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. In the study of nonlinear feedback systems they thus only apply in a limited regime where a first order approximation of the system is sufficiently exact.

\subsubsection{Error Signals and Controllers of Neural Homeostasis}
A key step in characterizing a control loop in the context of biological systems should be to identify the previously described abstract elements. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and ``straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. The latter is---as stated earlier---an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a ``sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological quantities to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section one could then ask what type of control best resembles the actual neural control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control, and rather points to an integral or proportional control. It is clear however that the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. The conductance of an ion channel for example has natural upper and lower bounds that it can not surpass. We will come back to this issue in Section \ref{Possible_Modifications_Section}.

\subsubsection{Single-Cell Homeostasis As an Integral Control}
Former versions of the LIF-SORN and SORN used a simple form of integral control to regulate excitatory firing rates. Individual firing rates of excitatory cells were estimated by a spike counter over short subsequent time periods and compared to a target value. This difference was then fed to a controller which increased or decreased neuronal excitability in an additive way (making it a discrete form of integral control). As mentioned before, this type of control allowed to reliably tune each cell to a given target firing rate under changing input. A global target firing rate was chosen for all excitatory neurons. Consequently, this resulted in a very narrow distribution of excitatory firing rates.

\subsubsection{The Role of Firing Rate Heterogeneity---a Dichotomy Between\\ Heterogeneity and Control?}\label{Section_Role_Fir_Rate_Het}
Modeling homeostasis necessarily includes a form of error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. Of course this raises the question of how to choose these homeostatic targets. As mentioned in the previous section, former versions of the SORN defined a single global target value. However, experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. Many statistical analyses found particular evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, claim that it is reasonable to define a single global target of activity.

However, numerous experimental and theoretical studies have suggested that the presence of both slow- and fast-firing cells is not to be regarded as an ignorable, mere side effect of brain dynamics \cite{Buzsaki_2004,Marsat_2010,Tripathy_2013}. It rather seems to be that skewed and heavy-tailed firing rate statistics are necessary with respect to the preservation of stable synaptic structures, being related to long-term memory, while still allowing synaptic rewiring in order to adapt to changes in external stimuli. For example, experiments by Dragoi et al. have found that during remapping of hippocampal place cells, neurons with low firing rates are subject to stronger changes within their field of spatial representation than those with high firing rates \cite{Dragoi_2003}. In this context Buzsáki and Mizuseki argue that a small subset of highly active neurons may serve to form relatively strongly connected and stable sub-networks that can quickly react to momentary changes in the environment, while fine tuned adaptation involves the slower firing majority of cells \cite{Buzsaki_Fir_Rates_2014}. A realistic network model of neuronal activity and plasticity should therefore allow for a broad distribution of firing rates to cope with these functional aspects.

@@

\subsubsection{Nitric Oxide in the Nervous System}\label{NO_Experiments_Section}
The model presented in \cite{Sweeney_Paper} includes spatial interaction across excitatory neurons through the diffusion of NO. Enzymes that are responsible for NO synthesis (NOS) are present in different areas of the body, namely being involved in the dilation of blood vessels (endothelial NOS, eNOS), the immune system (inducible NOS, iNOS) and the CNS (neural NOS, nNOS) \cite{NOS_Mammals}. These Enzymes differ in their functionality and dependence on the presence of other molecules. In particular, nNOS is sensitive to the concentration of $\mathrm{Ca^{2+}}$ \cite{Knowles_Ca_nNOS,Steinert_NO}. Nitric oxide that was synthesized in the nervous system can diffuse freely because of its small molecular structure and electrical neutrality, which makes it a good candidate to be involved into a form of diffusive intercellular communication \cite{Lancaster_1994}.

@@

\subsection{Previous Research on the Used Network Model}
This thesis is based on earlier versions of a Self-Organizing Recurrent Neural Network (SORN), in particular an implementation using leaky integrate-and-fire neurons (LIF-SORN) \cite{SORN_Paper}. It was developed to model features of activity and topology of cortical networks, in particular of L5 of the neocortex. A slow homeostatic mechanism was set up among excitatory neurons through an intrinsic adaptation of excitability in these previous implementations. This led to the aforementioned problems of homeostatic control: Neuronal activity was tuned to a globally set target, allowing for no variability. Despite this limitation, the model has proven to be capable of showing a number of experimentally confirmed features. Zheng et al. have shown that the distribution and dynamics of synaptic efficacies measured in rat hippocampus can be reproduced by a binary SORN using a discretized version of spike-timing-dependent plasticity and presynaptic normalization \cite{Pengsheng_2013}. Miner et al. designed the LIF-SORN based on this model. This network reproduced the results of the binary SORN regarding self-organizing synaptic dynamics. Furthermore, it was able to explain the experimentally observed overrepresentation of bidirectional connections (see \cite{Markram_Connections_1997,Song_Connectivity_2005}) by means of a distance-dependent implementation of synaptic growth.

@@

\subsection{Thesis Outline}
Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neuronal activity which has previously been suppressed by single-cell homeostasis. Furthermore, we predict excitatory activity on the level of individual cells by an analytic approach, thereby gaining an understanding of the relation between spatial structure and firing rates. In the face of possible new features within the network's structure, we clarify the causal relation between diffusive spatial interaction and synaptic topology.

@@

The neural network was simulated with the code used in \cite{SORN_Paper}, which makes use of the BRIAN spiking neural network simulator \cite{Briansim}. All following explanations regarding the simulation of neurons and mechanisms of synaptic plasticity are thus based on the methods described in the aforementioned paper.

400 excitatory LIF neurons and 80 inhibitory LIF neurons were assigned random positions across a square area of 1000 $\times$ 1000 $\rm \upmu m$. All except recurrent excitatory synapses were randomly generated before the start of the simulation until a desired connection fraction was reached. The connection probability between two neurons was calculated from a distance-dependent Gaussian function with a standard deviation of 200 $\rm \upmu m$. For excitatory to inhibitory (EI) and inhibitory to excitatory (IE) synapses, the connection fraction was set to 0.1, and 0.5 for recurrent inhibitory synapses (II). These connections were kept at a fixed connection strength throughout the simulation. Furthermore, all synapses were simulated with a fixed (distance independent) conduction delay. See table \ref{syn_conn_params} for a summary of parameters.

@@

\begin{table}
\caption{Parameters of synaptic connections.}
\begin{center}
\begin{tabu}{|l|l|l|l|l|}
\hline
\textbf{parameter} & \textbf{EE} & \textbf{EI} & \textbf{IE} & \textbf{II} \\ \hline
connection fraction & $\rightarrow 0.1$ & $0.1$ & $0.1$ & $0.5$ \\ \hline
initial connection strength & $\rm 0.0001\, mV$ & $\rm 1.5\, mV$ & $\rm -1.5\, mV$ & $\rm -1.5\, mV$ \\ \hline
conduction delay & $\rm 1.5\, ms$ & $\rm 0.5\, ms$ & $\rm 1.0\, ms$ & $\rm 1.0\, ms$ \\
\hline
\end{tabu}
\end{center}
\label{syn_conn_params}
\end{table}
